---
title: "HDRC_DataScientist"
author: "Chantel Davies"
date: "2023-12-05"
output:
  pdf_document: default
---
```{r setup, include=FALSE}
# Load packages
knitr::opts_chunk$set(echo = TRUE)
library(plyr)
library(tidyverse)
library(reactable)
library(gt)
library(readr)
library(purrr)
library(ggplot2)
library(sf)
library(leaflet)
library(htmltools)

sysfonts::font_add_google("Roboto")
```
# Introduction

The Director of Neighbourhoods wants to understand more about road traffic crashes in Bradford. You, the HDRC data scientist, have been asked to give some insight into road traffic crashes in Bradford.

Therefore, your task is to:
1) examine this Bradford road traffic crash data;
2) identify and display insights; and
3) give us a 10 minute presentation during your interview about the work you have done – this should include slides that showcase your work

# Questions/ areas of interest
Casualty 'hotspots'
Volume of casualties from 2018 to 2020 - how do they compare
Different types of casualties
Pedestrians and cyclists
Summary statistics (count, percentage initially)

```{r load-data, include=FALSE}
# Load crashes data
crashes <- read_csv("./data/bfd_crashes.csv",
                    col_names = TRUE)

# For some reason, the csv files wouldn't batch read, so I did it individually
# Subset to variables of interest
crashes <- crashes %>% select(c("accident_index",
                     "accident_year",
                     "longitude",
                     "latitude",
                     "number_of_casualties",
                     "date",
                     "day_of_week",
                     "time",
                     "local_authority_ons_district",
                     "road_type",
                     "speed_limit",
                     "pedestrian_crossing_human_control",
                     "pedestrian_crossing_physical_facilities",
                     "light_conditions"))

# Load vehicles data
vehicles <- read_csv("./data/bfd_vehicles.csv",
                     col_names = TRUE)

vehicles <- vehicles %>% select(c("accident_index",
                      "vehicle_type",
                      "sex_of_driver",
                      "age_band_of_driver",
                      "driver_imd_decile"))

#Load casualties data
casualties <- read_csv("./data/bfd_casualties.csv",
                       col_names = TRUE)

casualties <- casualties %>% select(c("accident_index",
                        "casualty_class",
                        "casualty_type",
                        "sex_of_casualty",
                        "age_band_of_casualty",
                        "casualty_severity",
                        "pedestrian_location",
                        "casualty_imd_decile"))

### Map ###
# I'll make a map and use the following shapefile
# https://ckan.publishing.service.gov.uk/dataset/local-authority-districts-may-2021-uk-bgc/resource/9fe4a677-79f9-4438-89e4-0f0370378ac3

# Read in with sf package
uk <- st_read(dsn = "data/Local_Authority_Districts_(May_2021)_UK_BGC/LAD_MAY_2021_UK_BGC.shp")
```


```{r basic-statistics, include=FALSE}
# Left join the data to match vehicles and crashes to casualties
bradford_rta <- join_all(list(casualties, vehicles, crashes), by='accident_index', type='left')

# Focus on the Bradford area using district code E08000032, as per the assignment specification
# Create a new 'month' column, position after the date, convert to factor and reorder the levels to chronological order
bradford_rta <- bradford_rta %>% 
  filter(local_authority_ons_district == "E08000032") %>% 
  mutate(month = format(as.Date(date, format="%Y-%m-%d"),"%B"), .after = date) %>%
  mutate(month = factor(month,
                        levels = c("January",
                                   "February",
                                   "March",
                                   "April",
                                   "May",
                                   "June",
                                   "July",
                                   "August",
                                   "September",
                                   "October",
                                   "November",
                                   "December")))


# A little function to quickly calculate percentages from the counts of different columns, from here: https://stackoverflow.com/questions/9623763/extend-contigency-table-with-proportions-percentages
# Useful for quickly eyeballing general counts and percentages

tblFun <- function(x){
    tbl <- table(x)
    res <- cbind(tbl,round(prop.table(tbl)*100,2))
    colnames(res) <- c('Count','Percentage')
    res
}

do.call(rbind, lapply(bradford_rta[3], tblFun))

#The greatest percentage of casualties overall were car occupants (67%), pedestrians (11%) and cyclists (9%) 

# Tally the number of casualties for each year to compare numbers
bradford_rta %>% 
  group_by(accident_year) %>% 
  summarise(total = n())

# The number of casualties declined from 2,432 in 2018 to 1,754 in 2020; 2020's numbers are likely to be a result of lockdowns and fewer cars on the road.

# Tally the number of casualties for each month of each year to look for any seasonal patterns and plot them on a line graph to compare
total_casualties <- bradford_rta %>% 
  group_by(accident_year, month) %>% 
  count()

total_casualties %>% 
  ggplot(aes(x = month, y = n, group = accident_year, color = as.factor(accident_year), shape = as.factor(accident_year))) +
  geom_line(linewidth = 1) +
  geom_point(size = 4) +
  labs(x = "Month",
       y = "Number of casualties") +
  scale_shape_manual(name = "Year",
                      values=c(15, 16, 17)) + 
  scale_color_manual(name = "Year",
                      values=c('#999999','#E69F00', '#56B4E9')) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 300)) +
  theme_bw() +
  theme(legend.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 14, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1))

```

There are casualty 'peaks' in October which is not uncommon and may be due to the switch from BST to GMT.

```{r drilling-down, include=FALSE}
# Going back to the three original datasets to produce a summary of volume and percentage of casualties and vehicle types for cyclists and pedestrians

bradford_rta %>% 
  group_by(casualty_type, vehicle_type) %>% 
  filter(casualty_type %in% c ("Cyclist", "Pedestrian")) %>% 
  summarise(n = n()) %>% 
  mutate(percentage = scales::percent(n/sum(n), accuracy = 1)) %>% 
  pivot_wider(names_from = casualty_type,
              values_from = c(n, percentage)) %>%
  select(vehicle_type, percentage_Cyclist, percentage_Pedestrian) %>% 
reactable(
    pagination = FALSE,
    sortable = TRUE,
    defaultColDef = colDef(
      align = "center"
    ),
    columns = list(
      vehicle_type = colDef(
        name = "Vehicle Type",
        align = "left"
      ),
      percentage_Cyclist = colDef(
        name = "% cyclists"
      ),
      percentage_Pedestrian = colDef(
        name = "% pedestrians"
      )
    ),
    defaultPageSize = 13) %>% 
  google_font(font_family = "Roboto")

```

85% of pedestrian casualties involve a car; 44% involve a car, but 50% involve a pedal cycle... which is sort of obvious, but may also be attributable to potholes, uneven road surfaces or other types of incidents - would be interesting to find out.

Let's have a look at the categories of casualty_severity for each time period

```{r further-exploration-code, include=FALSE}
# This shows the number of casualties by severity
# Initially I was only going to look at serious or fatal, but then went for all types as the slight casualties may lead to serious or fatal if not managed
bradford_rta %>% 
  #filter(casualty_severity %in% c("Serious", "Fatal")) %>% 
  group_by(casualty_severity, casualty_type) %>% 
  summarise(n = n()) %>% 
  print(n = Inf)

# Illustrated in a horizontal bar chart
bradford_rta %>% 
  #filter(casualty_severity %in% c("Serious", "Fatal")) %>% 
  group_by(casualty_severity, casualty_type) %>% 
  summarise(n = n()) %>% 
  ggplot(aes(x = casualty_type, y = n)) +
  geom_bar(stat='identity',
           width = 0.8) +
  geom_text(aes(label=n),
            hjust = -0.15) +
  coord_flip() +
  labs(y = "Number of casualties") +
  facet_grid(~casualty_severity) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 25)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title = element_blank(),
        axis.title.x = element_text(size = 16),
        axis.text = element_text(size = 8, colour = "black"),
        strip.background = element_rect(fill = "white"),
        strip.text = element_text(size = 12))

# Might need to zoom to get the full image

# Physical activity is important, so I will narrow down the type of casualty to pedestrians and cyclists
# Illustrated in a horizontal bar chart

seriousness <- bradford_rta %>% 
  filter(casualty_type %in% c("Cyclist", "Pedestrian")) %>% 
  group_by(casualty_severity, casualty_type) %>% 
  summarise(n = n())

seriousness %>% 
           ggplot(aes(x = casualty_type,
             y = n,,
             # fill = casualty_severity,
             text = paste("</br>Severity: ", casualty_severity,
                          "</br>Type: ", casualty_type,
                          "</br>Casualties: ", n))) +
  geom_bar(stat='identity',
           width = 0.8) +
  # coord_flip() +
  labs(x = "Casualty type",
       y = "Number of casualties") +
  facet_grid(~casualty_severity) +
  #scale_x_discrete(labels = function(x) str_wrap(x, width = 25)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14, vjust = -1),
        axis.text = element_text(size = 8, colour = "black"),
        strip.background = element_rect(fill = "white"),
        strip.text = element_text(size = 14))
```

Pedestrians and cyclists are not faring too well on Bradford's roads...

With the magic of {leaflet}, we can see where the hotspots occur for subgroups of pedestrians and cyclists.

```{r map, include=FALSE}

# Subsetting the geometry of the shapefile imported earlier
# Filter for Bradford and removed unneeded variables
bradford <- uk %>% 
  filter(LAD21CD == "E08000032") %>% 
  select(-c(OBJECTID, LAD21NM, LAD21NMW, BNG_E, BNG_N, GlobalID, LONG, LAT))

# Reduce the size of the shapefile to make R less likely to crash
bradford <- st_simplify(bradford, preserveTopology = TRUE, dTolerance = 1000)

# Remove the larger shapefile
rm(uk)

# I want to show accident hotspots on the map, so I combine the accident data with the shapefile - don't have to do it this way, but it can be helfpul in the eventualty of using packages like crosstalk
bradford_data <- left_join(bradford_rta, bradford, by = c("local_authority_ons_district" = "LAD21CD"))

# Convert to sf format
bradford_data <- st_as_sf(bradford_data)

# Transform the dataframe to a geometry object with the appropriate projection
bradford_data <- st_transform(bradford_data, '+proj=longlat +datum=WGS84')

# Subset to variables and levels of interest
bradford_mapping <- bradford_data %>% 
  filter(casualty_type %in% c("Cyclist", "Pedestrian") & casualty_severity %in% c("Fatal", "Serious")) %>% 
  mutate(casualty_type = factor(casualty_type))

# Colour coding the points to indicate whether the casualty is a pedestrian or cyclist
pal <- colorFactor(palette = c("darkblue", "steelblue"), domain = bradford_mapping$casualty_type)

# Onto the map
leaflet(bradford_mapping) %>% 
   setView(lng = -1.82, lat = 53.81, zoom = 11.5) %>%
   addProviderTiles(providers$Esri.WorldStreetMap) %>% 
   addCircleMarkers(lng = ~longitude,
                    lat = ~latitude,
                    popup = ~paste("Type: ", casualty_type,
                                   "</br>Severity: ", casualty_severity,
                                   "</br>", month, " ", accident_year,
                                   "</br>Speed Limit: ", speed_limit,
                                   "</br>Crossing: ", pedestrian_crossing_physical_facilities),
                    stroke = FALSE,
                    fillOpacity = 0.75,
                    color = ~pal(casualty_type),
                    clusterOptions = markerClusterOptions(palette())) %>% 
  addLegend(pal = pal,
            values = ~casualty_type,
            title = "Type of casualty (pedestrian or cyclist)",
            opacity = 1)

```


```{r summary-table, include=FALSE}
# This table is very useful for filtering data and learning more about what is happening to pedestrians and cyclists within the severe or fatal casualty category

# Run the code and play with the table
reactable(bradford_mapping %>% dplyr::select(accident_year, casualty_type, casualty_severity, sex_of_casualty, age_band_of_casualty, pedestrian_crossing_human_control, speed_limit),
          height = 500,
          highlight = TRUE,
          wrap = TRUE,
          sortable = TRUE,
          resizable = TRUE,
          searchable = FALSE,
          groupBy = "casualty_type",
          defaultColDef = colDef(
            show = FALSE,
            minWidth = 70,
            filterable = TRUE,
            align = "left"
          ),
          columns = list(
            casualty_type = colDef(
              show = TRUE,
              name = "Type"
            ),
            accident_year = colDef(
              show = TRUE,
              name = "Year",
              maxWidth = 70
            ),
            sex_of_casualty = colDef(
              show = TRUE,
              name = "Sex",
              maxWidth = 70
            ),
            age_band_of_casualty = colDef(
              show = TRUE,
              name = "Age category",
              maxWidth = 130,
              align = "center"
            ),
            casualty_severity = colDef(
              show = TRUE,
              name = "Severity",
              filterable = TRUE,
              maxWidth = 200,
              align = "center"
            ),
            pedestrian_crossing_human_control = colDef(
              show = TRUE,
              name = "Pedestrian crossing",
              align="center"
            ),
            speed_limit = colDef(
              show = TRUE,
              name = "Speed limit",
              maxWidth = 100,
              align = "center"
            )
          ),
  defaultPageSize = 13) %>% 
  google_font(font_family = "Roboto")
```